---
title: "Aggregation/Scaling performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation/Scaling performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
Aggregations and scaling with `hierarchyUtils` should perform as fast as basic data.table code.
For basic use cases there should only be slightly more overhead due to the assertions and added flexibility included in `hierarchyUtils`.

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  message = FALSE,
  comment = "#>"
)
```

```{r setup}
library(hierarchyUtils)
library(data.table)

profile_n_draws <- c(1, 10, 100, 1000)
```

```{r time_aggregation_function}
#' @title Helper function to time `hierarchyUtils::agg` and basic data.table
#'   aggregation.
#'
#' @param n_draws \[`integer(1)`\]\cr
#'   Number of draws to expand `id_vars` by.
#' @param id_vars \[`list(1)`\]\cr
#'   List of variables to expand to use as test input dataset. `id_vars` is
#'   passed to `data.table::CJ`. Assumed to not include 'draw'. Assumed to include
#'   'age_start' & 'year_start' but not the end variables.
#' @inheritParams hierarchyUtils::agg
#'
#' @return \[`data.table(1)`\] containing summary information about the timing results.
time_aggregation <- function(n_draws,
                             id_vars,
                             col_stem,
                             col_type,
                             mapping) {
  
  # create input dataset
  id_vars <- copy(id_vars)
  id_vars[["draw"]] <- 1:n_draws
  input_dt <- do.call(CJ, id_vars)
  
  # add interval end columns
  input_dt[, year_end := year_start + 1]
  input_dt[, age_end := age_start + 1]
  input_dt[age_start == 95, age_end := Inf]
  
  # identify value and id cols
  value_cols <- grep("value", names(input_dt), value = TRUE)
  id_cols <- names(input_dt)[!names(input_dt) %in% value_cols]
  
  # time hierarchyUtils aggregation function
  start_time <- proc.time()
  hierarchyUtils_output_dt <- agg(
    dt = input_dt,
    id_cols = id_cols, value_cols = value_cols,
    col_stem = col_stem, col_type = col_type,
    mapping = mapping
  )
  end_time <- proc.time()
  hierarchyUtils_time <- end_time - start_time
  
  # time basic data.table aggregation
  start_time <- proc.time()
  
  cols <- col_stem
  if (col_type == "interval") {
    cols <- paste0(col_stem, "_", c("start", "end"))
  }
  by_id_cols <- id_cols[!id_cols %in% cols]

  if (col_type == "interval") {
    
    data.table_output_dt <- lapply(1:nrow(mapping), function(i) {
      start_bound <- mapping[i, get(cols[1])]
      end_bound <- mapping[i, get(cols[2])]
    
      agg_dt <- input_dt[
        get(cols[1]) >= start_bound & get(cols[2]) <= end_bound,
        lapply(.SD, sum), .SDcols = value_cols,
        by = by_id_cols
      ]
      agg_dt[, c(cols) := list(start_bound, end_bound)]
      return(agg_dt)
    })
    data.table_output_dt <- rbindlist(data.table_output_dt)
    
  } else {
    
    data.table_output_dt <- input_dt[
      get(col_stem) %in% mapping[, child],
      lapply(.SD, sum), .SDcols = value_cols,
      by = by_id_cols
    ]
    data.table_output_dt[, c(col_stem) := list(mapping[, unique(parent)])]
  }
  end_time <- proc.time()
  data.table_time <- end_time - start_time
  
  # check that outputs are the same
  setcolorder(hierarchyUtils_output_dt, c(id_cols, value_cols))
  setkeyv(hierarchyUtils_output_dt, id_cols)
  setcolorder(data.table_output_dt, c(id_cols, value_cols))
  setkeyv(data.table_output_dt, id_cols)
  testthat::expect_identical(hierarchyUtils_output_dt, data.table_output_dt)

  # compile together timings in formatted table
  timings <- lapply(c("hierarchyUtils", "data.table"), function(method) {
    elapsed_time <- get(paste0(method, "_time"))
    elapsed_time <- as.list(elapsed_time)
    setDT(elapsed_time)
    elapsed_time[, method := method]
    elapsed_time[, c("col_stem", "col_type") := list(col_stem, col_type)]
    elapsed_time[, n_draws := n_draws]
    elapsed_time[, n_input_rows := format(nrow(input_dt), big.mark = ",", scientific = FALSE)]
  })
  timings <- rbindlist(timings)
  return(timings)
}
```

```{r time_aggregations}
# default variables for aggregation timings
age_mapping <- data.table(age_start = c(0, seq(0, 90, 5)), age_end = c(Inf, seq(5, 95, 5)))
sex_mapping <- data.table(parent = "all", child = c("male", "female"))
agg_id_vars <- list(
  location = 1,
  year_start = seq(1950, 2020, 1),
  sex = c("male", "female"),
  age_start = seq(0, 95, 1),
  value1 = 1, value2 = 1
)

agg_timings <- lapply(profile_n_draws, function(n_draws) {
  draw_timings <- list(
    age_timing <- time_aggregation(
      n_draws = n_draws,
      id_vars = agg_id_vars,
      col_stem = "age", col_type = "interval",
      mapping = age_mapping
    ),
    sex_timing <- time_aggregation(
      n_draws = n_draws,
      id_vars = agg_id_vars,
      col_stem = "sex", col_type = "categorical",
      mapping = sex_mapping
    )
  )
  draw_timings <- rbindlist(draw_timings)
  return(draw_timings)
})
agg_timings <- rbindlist(agg_timings)
```

## Aggregation timing comparison

This vignette makes aggregation timing comparisons between `hierarchyUtils` and `data.table` for example input data when aggregating across an interval variable like 'age' or a categorical variable like 'sex'.
The example input data increases in size as more draws are added.

### Example input data

```{r example_input_data}
# create input dataset
agg_id_vars <- copy(agg_id_vars)
agg_id_vars[["draw"]] <- 1:1
input_dt <- do.call(CJ, agg_id_vars)
input_dt
```

### Timing comparison

```{r display_agg_timings, results="asis"}
setcolorder(agg_timings, c("col_stem", "col_type", "n_draws", "method", "n_input_rows"))
setkeyv(agg_timings, c("col_stem", "col_type", "n_draws", "method", "n_input_rows"))
agg_timings[, c("user.child", "sys.child") := NULL]
knitr::kable(agg_timings)
```
